---
title: "Exploratory Data Analysis on Hospital Billing Data"
date: "`r Sys.Date()`"
author: "Rui Qiu"
output:
  rmdformats::robobook:
    highlight: kate
    lightbox: true
---

```{r, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

```{r, echo=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, janitor, skimr, styler, ggtext, showtext,
  ggthemes, readxl, lubridate, glue, patchwork, MetBrewer,
  geofacet, tidytext, arrow
)

font_add_google(name = "Roboto Mono", family = "Roboto Mono")

theme_set(
  theme_minimal() +
    theme(
      legend.position = "bottom",
      legend.background = element_rect(fill = "#F9EFE6", color = "#F9EFE6"),
      legend.key = element_rect(fill = "#F9EFE6", color = "#F9EFE6"),
      legend.title = element_text(size = 7, color = "#3B372E"),
      legend.text = element_text(size = 7, color = "#3B372E"),
      plot.background = element_rect(fill = "#F9EFE6", color = "#F9EFE6"),
      panel.background = element_rect(fill = "#F9EFE6", color = "#F9EFE6"),
      text = element_text(
        family = "Roboto Mono",
        color = "#3B372E"
      ),
      axis.title.y = element_text(vjust = 0.2, face = "bold"),
      axis.title.x = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(),
      axis.text.y = element_text(angle = 30),
      plot.title = element_markdown(
        size = 18, hjust = 0.5,
        family = "Roboto Slab"
      ),
      plot.subtitle = element_markdown(
        hjust = 0.5,
        family = "Roboto Slab"
      ),
      plot.caption = element_markdown(
        size = 10,
        family = "Roboto Slab",
        hjust = 0
      ),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = margin(15, 15, 15, 15)
    )
)
```

# Preliminary cleaning

Manipulations:

- Remove rows with MRN only (NA in other columns).
- Mutate `pt_deceased` to a logical variable.
- Unify “unknown” categories in `pt_race`.
- Unify “unknown” categories in `pt_ethnicity`.

```{r, echo=FALSE}
data <- read_xlsx("data/Billing_Data_15_Mar_2022_deid2.xlsx",
  col_types = c(
    rep("guess", 33), rep("text", 4),
    rep("date", 2)
  )
) |>
  clean_names()

# skim(data)

# Starting from line 47636, the records contain nothing but MRN code, so we believe it's okay to get rid of that part for now

# One of the identifier (not recommended but effective here) is to use dept_name (which is doubled-checked with `sum(!is.na(data$dept_name))`)

# loc_cat_2_name is a good cateogry to use in fact

data <- data |>
  drop_na(dept_name) |>
  mutate(
    pt_deceased = if_else(
      pt_deceased == "N", FALSE, TRUE
    ),
    pt_race = if_else(
      pt_race %in% c(
        "DECLINED TO ANSWER",
        "DECLINED/UNKNOWN",
        "Unknown",
        "UNKNOWN"
      ),
      "DECLINED/UNKNOWN", pt_race
    ),
    pt_ethnicity = if_else(
      pt_ethnicity %in% c("UNKNOWN", "DECLINED TO ANSWER") |
        is.na(pt_ethnicity),
      "DECLINED/UNKNOWN", pt_ethnicity
    )
  )
```

After cleaning, `skim()` would reveal the general structure of the remaining data set:

```{r, echo=FALSE}
skim(data)
```

So basically, we have:

- 47634 valid records of hospital billings. Each record consists of 39 columns.
- All character variables excluding diagnoses are complete, without missing values.
  - Note that some categorical data are marked as "unknown".

Starting from here, we will group data by unique patient as some patients have multiple billing records.

# EDA

One thing to notice here is that the medical record number (MRN) should be unique for each patient, but there seems to be a discrepancy in Date of Birth:

```{r, echo=FALSE, message=FALSE}
data |>
  select(new_mrn, new_dob, pt_sex, pt_race, pt_deceased, pt_ethnicity) |>
  filter(new_mrn == "51587384")
```

## Patients

- Add a manual `ID` to distinguish each patient.

> If we randomly assign a `DOB` to the records under the same `MRN`, we could've ended up with some ages < 0 by the time of the visit.

- Calculate the age by the time of the service.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
data <- data |>
  mutate(
    ID = group_indices(data, new_mrn, new_dob),
    age = interval(new_dob, new_srvc_date) %/% years(1)
  ) |>
  relocate(ID)

# write_csv(data, "data/cleaned-data.csv")
# write_parquet(data, "data/cleaned-data.parquet")
save(data, file = "cleaned-data.rda")
```

```{r, echo=FALSE, message=FALSE}
grp_data <- data |>
  group_by(ID) |>
  summarize(
    age = mean(age),
    record_num = n(),
    pt_sex = pt_sex,
    pt_race = pt_race,
    pt_ethnicity = pt_ethnicity,
    pt_state = pt_state,
    pt_deceased = pt_deceased
  )

# write_csv(grp_data, "data/grouped-data.csv")
# write_parquet(grp_data, "data/grouped-data.parquet")
save(grp_data, file = "grouped-data.rda")
```

### Age distribution

```{r, echo=FALSE, message=FALSE}
grp_data |>
  drop_na(age) |>
  ggplot(aes(
    x = age,
    color = pt_sex,
    fill = pt_sex
  )) +
  geom_histogram(alpha = 0.5, bins = 50) +
  facet_wrap(. ~ pt_sex) +
  scale_colour_manual(
    values = met.brewer("Hiroshige", 2, "discrete"),
    aesthetics = c("fill", "color")
  ) +
  labs(
    title = glue("THE AGE DISTRIBUTION OF
                  PATIENTS"),
    subtitle = glue("<span style='color:#72BCD5;'>**Male**</span> and <span style='color:#EF8A47;'>**Female**</span> patients' age are calculated by the time of their visits."),
    caption = glue("**Source**: Hospital Billing Dataset<br>
                   **Plot by**: Rui Qiu"),
    x = "age",
    y = "patient count"
  ) +
  theme(
    legend.position = "none"
  )

ggsave(
  filename = "figures/figure1.png",
  plot = last_plot(),
  height = 8, width = 8
)
```

### Geolocation

```{r, echo=FALSE, message=FALSE}
grp_data |>
  group_by(pt_state, pt_race) |>
  summarize(count = n()) |>
  ggplot(aes(x = pt_race, y = count, fill = pt_race)) +
  geom_col() +
  coord_flip() +
  facet_geo(~pt_state) +
  scale_colour_manual(
    values = met.brewer("Hiroshige", 7, "discrete"),
    aesthetics = c("fill", "color")
  ) +
  scale_y_continuous(breaks = c(0, 4000, 8000)) +
  labs(
    title = glue("THE RACE DISTRIBUTION BY STATES"),
    caption = glue("**Source**: Hospital Billing Dataset<br>
                   **Plot by**: Rui Qiu"),
    x = "",
    y = "patient count"
  ) +
  theme(
    axis.title.y = element_text(vjust = 0.2, face = "bold"),
    axis.title.x = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 5, angle = 0)
  )

ggsave(
  filename = "figures/figure2.png",
  plot = last_plot(),
  height = 8, width = 12
)
```

**Click [here](figures/figure2.png) for better resolution and scale.**

### Time of visiting by major states

```{r, echo=FALSE, message=FALSE}
data |>
  filter(pt_state %in% c("VA", "MD", "DC")) |>
  select(new_srvc_date, pt_state) |>
  mutate(new_srvc_date = make_date(
    year = year(new_srvc_date),
    month = month(new_srvc_date)
  )) |>
  group_by(new_srvc_date, pt_state) |>
  summarise(count = n()) |>
  drop_na(new_srvc_date) |>
  ggplot(aes(x = new_srvc_date, y = count, color = pt_state)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 350)) +
  scale_colour_manual(
    values = met.brewer("Hiroshige", 3, "discrete"),
    aesthetics = c("fill", "color")
  ) +
  labs(
    title = glue("MONTHLY VISITS BY STATES"),
    subtitle = glue("Missing values excluded."),
    caption = glue("**Source**: Hospital Billing Dataset<br>
                   **Plot by**: Rui Qiu"),
    x = "date",
    y = "patient visits"
  )

ggsave(
  filename = "figures/figure3.png",
  plot = last_plot(),
  height = 8, width = 8
)
```

## Diagnosis

Due to limited prior medical knowledge, I only managed to combine multiple diagnoses of the same patient together to see if some keywords are predominant in the large picture.

> A better approach would be using some pre-defined medical term dictionary to match the patterns of diagnoses.

The resulted data frame looks like this:

```{r, echo=FALSE, message=FALSE}
tmp <- data |>
  select(ID, starts_with("dx") & ends_with("desc")) |>
  mutate(dx = paste(
    dx1_desc, dx2_desc, dx3_desc,
    dx4_desc, dx5_desc, dx6_desc,
    dx7_desc, dx8_desc, dx9_desc
  )) |>
  select(ID, dx) |>
  mutate(dx = str_replace_all(dx, " NA | NA|NA ", "")) |>
  distinct()

tmp |>
  top_n(10) |>
  knitr::kable()
```

```{r, echo=FALSE, message=FALSE}
tmp |>
  unnest_tokens(word, dx) |>
  anti_join(stop_words) |>
  count(word, sort = TRUE) |>
  top_n(20) |>
  mutate(word = reorder(word, n)) |>
  ggplot(aes(n, word)) +
  geom_col() +
  scale_colour_manual(
    values = met.brewer("Hiroshige", 1, "discrete"),
    aesthetics = c("fill", "color")
  ) +
  labs(
    title = glue("FREQUENT MEDICAL TERMINOLOGIES"),
    sutitle = glue("IN DIAGNOSES"),
    caption = glue("**Source**: Hospital Billing Dataset<br>
                   **Plot by**: Rui Qiu"),
    x = "word count",
    y = "terminology"
  )

ggsave(
  filename = "figures/figure4.png",
  plot = last_plot(),
  height = 8, width = 8
)
```

# Dashboard

Another page of Shiny dashboard aims to tackle the following tasks.

It is accessible at [this link](https://rexarski.shinyapps.io/hospital-billing/)


```{r, echo=FALSE, message=FALSE}
# # Extract the scripts used in this file as a separate .R file
# knitr::purl(
#   input = "index.Rmd",
#   output = "script.R",
#   documentation = 0
# )
#
# style_file("index.Rmd")
```
